When environmentally persistent pathogens transform good habitat into ecological traps

### 2016-01-21 Manuscript/manuscript.tex
Additions to future work and preference discussion

Thus, if high quality habitat is preferred by migrants (i.e. $\xi_{em} > 0$), it forms an ecological trap at the individual scale, that is, preferentially selected habitat that reduces individual fitness relative to other habitats (\cite{Robertson2006}).
Moreover, host preference for high quality habitat further exacerbates the increase in disease risk on high quality habitat (Supplement Figs 5, 7) by increasing the rate of pathogen import from infectious migrants also selecting high quality habitat.
This echoes results from the contact network literature that show that highly connected nodes have higher risk of becoming infected (\cite{Christley2005}, \cite{Keeling2005}).

@@

The framework used here provides general insights, but generating predictions for specific systems requires a careful consideration of the model's simplifying assumptions.
In particular, the stochastic patch occupancy model, though well suited for capturing realistic spatial structure, assumes that within-patch dynamics (i.e. growth to carrying capacity and local pathogen spread) are rapid relative to among-patch dynamics (i.e. colonization and extinction).
This is often a reasonable assumption, especially for relatively small populations in a highly fragmented landscape (\cite{Hanski2003}), but may not capture the non-linearities and nuances of host-pathogen dynamics in a single population.
In addition, we assume that patch quality only influences population size and the resulting colonization and extinction processes, but quality could also influence pathogen dynamics within a patch through effects on host density or condition (\cite{Becker2015}).
Patch quality may also influence the size or infectiousness of the environmental reservoir, either directly, by influencing the decay rate of the pathogen in the environment, or indirectly, through the effects on population size and overall shedding rate.
The potential importance of these additional mechanisms remains an open question for future work and needs to be considered before applying this framework to any given system of interest.
Fortunately, the SPOM framework is fairly flexible and future work could extend the model presented here to include additional processes, e.g. by making transmission parameters also function of patch quality.

>>>>

Thus, if high quality habitat is preferred by migrants (i.e. $\xi_{em} > 0$), it forms an ecological trap at the individual scale, that is, preferentially selected habitat that reduces individual fitness relative to other habitats (\cite{Robertson2006}).
Moreover, host preference for high quality habitat further exacerbates this increase in disease risk (Supplement Figs 5, 7) by increasing the rate of pathogen import from infectious migrants also selecting high quality habitat.
The magnitude of this effect depends on the landscape structure and the scale at which individuals sense and select habitat.
In a fully connected metapopulation, where high quality patches attract migrants from the entire landscape, the effect of preference on the relationship between quality and disease risk is the most pronounced (Supplement Fig 7).
However, the fact that the effect persists in a lattice metapopulation (Supplement Fig 5), where migrants can only select among the four neighboring patches, suggests that host preference for high quality habitat likely serves to further increase the disease risk on high quality patches even in more realistic landscapes.
This echoes results from the contact network literature that show that highly connected nodes have higher risk of becoming infected (\cite{Christley2005}, \cite{Keeling2005}).

@@

The framework used here provides general insights, but generating predictions for specific systems requires a careful consideration of the model's simplifying assumptions.
In particular, the stochastic patch occupancy model, though well suited for capturing realistic spatial structure, assumes that within-patch dynamics (i.e. growth to carrying capacity and local pathogen spread) are rapid relative to among-patch dynamics (i.e. colonization and extinction).
This is often a reasonable assumption, especially for relatively small populations in a highly fragmented landscape (\cite{Hanski2003}), but may not capture the non-linearities and nuances of host-pathogen dynamics in a single population.
In addition, we assume that patch quality only influences population size and the resulting colonization and extinction processes, but quality could also influence pathogen dynamics within a patch through effects on host density or condition (\cite{Becker2015}).
Patch quality may also influence the size or infectiousness of the environmental reservoir, either directly, by influencing the decay rate of the pathogen in the environment, or indirectly, through the effects on population size and overall shedding rate.
More generally, the dynamics and establishment of environmental reservoirs could be influenced by a number of other factors, such as the length of infectious occupancy or the occupancy history of a patch.
The potential importance of these additional mechanisms remains an open question for future work and needs to be considered before applying this framework to any given system of interest.
Fortunately, the SPOM framework is fairly flexible and future work could extend the model presented here to include additional processes, e.g. by making transmission parameters also function of patch quality.
